---
title: "Final_Project_EDA"
author: "Anna Hyunjung Kim"
date: "2026-04-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```


```{r}
student <- read.csv("./student/student-por.csv", header = TRUE, sep = ";")
```
# Variables Table

## Student Performance Dataset (UCI) – Variable Description

Dataset: Student Performance Dataset (https://archive.ics.uci.edu/dataset/320/student+performance)

This dataset contains information about secondary school students, including demographic characteristics, family background, academic behavior, and social factors. It is commonly used to analyze and predict student academic performance.


---

### Target Variable

| Variable | Description |
|----------|-------------|
| G3 | Final grade (0–20)|
| G2 | Second period grade |
| G1 | First period grade |


---

### Demographic Variables

| Variable | Description |
|----------|-------------|
| school | School attended (GP: Gabriel Pereira, MS: Mousinho da Silveira) |
| sex | Student’s sex (F: female, M: male) |
| age | Student’s age (15–22) |
| address | Home address type (U: urban, R: rural) |
| famsize | Family size (LE3: ≤3 members, GT3: >3 members) |
| Pstatus | Parent cohabitation status (T: together, A: apart) |

---

### Family Background

| Variable | Description |
|----------|-------------|
| Medu | Mother’s education level (0–4) |
| Fedu | Father’s education level (0–4) |
| Mjob | Mother’s job |
| Fjob | Father’s job |
| guardian | Student’s guardian (mother, father, or other) |
| famsup | Family educational support (yes/no) |

---

### School and Study-related Variables

| Variable | Description |
|----------|-------------|
| traveltime | Travel time to school (1: <15 min, 4: >1 hour) |
| studytime | Weekly study time (1: <2 hours, 4: >10 hours) |
| failures | Number of past class failures |
| schoolsup | Extra educational support from school (yes/no) |
| paid | Extra paid classes (yes/no) |
| activities | Participation in extracurricular activities (yes/no) |
| nursery | Attended nursery school (yes/no) |
| higher | Intention to pursue higher education (yes/no) |

---

### Social and Lifestyle Variables

| Variable | Description |
|----------|-------------|
| internet | Internet access at home (yes/no) |
| romantic | In a romantic relationship (yes/no) |
| famrel | Quality of family relationships (1–5) |
| freetime | Free time after school (1–5) |
| goout | Going out with friends (1–5) |
| Dalc | Workday alcohol consumption (1–5) |
| Walc | Weekend alcohol consumption (1–5) |
| health | Current health status (1–5) |

---

# Exploratory Data Analysis (EDA)

```{r}
colSums(is.na(student))
```


```{r}
head(student)
```


```{r}
str(student)
```

```{r}
dim(student)
```

### Distribution of Target Variables
```{r}
par(mfrow = c(1, 3))

hist(student$G1, main = "Distribution of Final Grade (G1)", xlab = "G1")
hist(student$G2, main = "Distribution of Final Grade (G2)", xlab = "G2")    
hist(student$G3, main = "Distribution of Final Grade (G3)", xlab = "G3")
```

```{r}
boxplot(student$G3, horizontal = TRUE)
```

The pattern of G1, G2, and G3 look similar. So, we need to check the correlation. 

### Relationship Between G1, G2, and G3

```{r}
plot(student$G1, student$G3, main = "G1 vs G3", xlab = "G1", ylab = "G3")
plot(student$G2, student$G3, main = "G2 vs G3", xlab = "G2", ylab = "G3")
```

```{r}
cor(student$G1, student$G3)
cor(student$G2, student$G3)
```

The correlations are very high (0.83 and 0.92), indicating a strong linear relationship. Therefore, G3 is sufficient as the final outcome variable.

### Demographic Variables

```{r}
library(ggplot2)
library(gridExtra)
```


```{r}
p1 <- ggplot(student, aes(x = school, y = G3)) + geom_boxplot() + ggtitle("School")
p2 <- ggplot(student, aes(x = sex, y = G3)) + geom_boxplot() + ggtitle("Sex")
p3 <- ggplot(student, aes(x = factor(age), y = G3)) + geom_boxplot() + ggtitle("Age")
p4 <- ggplot(student, aes(x = address, y = G3)) + geom_boxplot() + ggtitle("Address")
p5 <- ggplot(student, aes(x = famsize, y = G3)) + geom_boxplot() + ggtitle("Family Size")
p6 <- ggplot(student, aes(x = Pstatus, y = G3)) + geom_boxplot() + ggtitle("Parent Status")

grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 3)
```

- School: GP students score slightly higher than MS students.
- Sex: Female and male students have similar grades. Not a big difference.
- Age: Younger students (15–16) tend to score higher. Older students do worse.
- Address: Urban students score a little higher than rural students.
- Family Size & Parent Status: Almost no difference. These may not be important predictors.

### Family Background
```{r}
p1 <- ggplot(student, aes(x = factor(Medu), y = G3)) + geom_boxplot() + ggtitle("Mother's Education")
p2 <- ggplot(student, aes(x = factor(Fedu), y = G3)) + geom_boxplot() + ggtitle("Father's Education")
p3 <- ggplot(student, aes(x = Mjob, y = G3)) + geom_boxplot() + ggtitle("Mother's Job") + theme(axis.text.x = element_text(angle = 45, hjust = 1))
p4 <- ggplot(student, aes(x = Fjob, y = G3)) + geom_boxplot() + ggtitle("Father's Job") + theme(axis.text.x = element_text(angle = 45, hjust = 1))
p5 <- ggplot(student, aes(x = guardian, y = G3)) + geom_boxplot() + ggtitle("Guardian")
p6 <- ggplot(student, aes(x = famsup, y = G3)) + geom_boxplot() + ggtitle("Family Support")

grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 3)
```

- Mother's & Father's Education: Higher parental education -> higher student grades. Clear trend!
- Mother's Job: Students whose mothers work in health or teaching score higher.
- Father's Job: Similar pattern. Teacher/health jobs -> better grades.
- Guardian: Not much difference between mother, father, or other.
- Family Support: Interestingly, students without family support score slightly higher. Maybe those students are more independent.



### School and Study-related Variables

```{r}
p1 <- ggplot(student, aes(x = factor(traveltime), y = G3)) + geom_boxplot() + ggtitle("Travel Time")
p2 <- ggplot(student, aes(x = factor(studytime), y = G3)) + geom_boxplot() + ggtitle("Study Time")
p3 <- ggplot(student, aes(x = factor(failures), y = G3)) + geom_boxplot() + ggtitle("Failures")
p4 <- ggplot(student, aes(x = schoolsup, y = G3)) + geom_boxplot() + ggtitle("School Support")
p5 <- ggplot(student, aes(x = paid, y = G3)) + geom_boxplot() + ggtitle("Paid Classes")
p6 <- ggplot(student, aes(x = activities, y = G3)) + geom_boxplot() + ggtitle("Activities") 

grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 3)
```

```{r}
p7 <- ggplot(student, aes(x = nursery, y = G3)) + geom_boxplot() + ggtitle("Nursery")
p8 <- ggplot(student, aes(x = higher, y = G3)) + geom_boxplot() + ggtitle("Higher Education")
grid.arrange(p7, p8, ncol = 2)
```
- Study Time: More study time -> higher grades. Makes sense!
- Failures: Students with more past failures score much lower. This is a very strong predictor.
- School Support: Students with school support score lower — maybe because struggling students get support.
- Paid Classes: Little difference.
- Activities: Little difference.
- Higher Education (goal): Students who want to go to university score much higher. Big difference!


### Social and Lifestyle Variables

```{r}
p1 <- ggplot(student, aes(x = internet, y = G3)) + geom_boxplot() + ggtitle("Internet")
p2 <- ggplot(student, aes(x = romantic, y = G3)) + geom_boxplot() + ggtitle("Romantic")
p3 <- ggplot(student, aes(x = factor(famrel), y = G3)) + geom_boxplot() + ggtitle("Family Relationship")
p4 <- ggplot(student, aes(x = factor(freetime), y = G3)) + geom_boxplot() + ggtitle("Free Time")
p5 <- ggplot(student, aes(x = factor(goout), y = G3)) + geom_boxplot() + ggtitle("Going Out")
p6 <- ggplot(student, aes(x = factor(Dalc), y = G3)) + geom_boxplot() + ggtitle("Workday Alcohol")

grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 3)
```


```{r}
p7 <- ggplot(student, aes(x = factor(Walc), y = G3)) + geom_boxplot() + ggtitle("Weekend Alcohol")
p8 <- ggplot(student, aes(x = factor(health), y = G3)) + geom_boxplot() + ggtitle("Health")

grid.arrange(p7, p8, ncol = 2)
```

- Internet: Students with internet access score slightly higher.
- Romantic: Little difference.
- Family Relationship: Not a strong trend.
- Free Time & Going Out: More free time or going out -> slightly lower grades, but not dramatic.
- Workday & Weekend Alcohol: Higher alcohol use -> slightly lower grades. Weak but noticeable trend.
- Health: No clear pattern. Health does not seem to predict grades well.

## numeric variables correlation

```{r}
num_vars <- student[, c("G1","G2","G3","studytime","failures","absences","famrel","freetime","goout","Dalc","Walc","health")]
cor_matrix <- cor(num_vars)
cor_matrix
```
```{r}
library(corrplot)
```


```{r}
#install.packages("ggcorrplot")
library(ggcorrplot)

ggcorrplot(cor_matrix, lab = TRUE, lab_size = 3)
```

















