---
title: "KNN"
author: "Aditya Kataria"
date: "2026-04-18"
output: html_document
---

```{r}
student <- read.csv("./student/student-por.csv", sep = ";")
```


```{r}
library(FNN)

# same predictors as BIC model 
predictors <- c("school", "Fedu", "studytime", "failures", "schoolsup", 
                "higher", "Dalc", "health")

# Encode factors as numeric
df_knn <- student[, c(predictors, "G3")]
df_knn$school <- ifelse(df_knn$school == "GP", 0, 1)
df_knn$schoolsup <- ifelse(df_knn$schoolsup == "yes", 1, 0)
df_knn$higher <- ifelse(df_knn$higher == "yes", 1, 0)
df_knn <- na.omit(df_knn)

nrow(df_knn)

# Train/test split (80/20)
set.seed(1)
train_idx <- sample(1:nrow(df_knn), 0.8 * nrow(df_knn))
train <- df_knn[train_idx, ]
test  <- df_knn[-train_idx, ]

train_scaled <- scale(train[, predictors])
train_center <- attr(train_scaled, "scaled:center")
train_scale  <- attr(train_scaled, "scaled:scale")

train_X <- as.matrix(train_scaled)
test_X  <- as.matrix(scale(test[, predictors],
                            center = train_center,
                            scale  = train_scale))
train_Y <- train$G3
test_Y  <- test$G3

# Find optimal K
K <- 1:50
mse_vals <- numeric(50)
for (k in K) {
  pred <- knn.reg(train_X, test_X, train_Y, k = k)$pred
  mse_vals[k] <- mean((pred - test_Y)^2)
}

best_k <- which.min(mse_vals)
best_k
mse_vals[best_k]

plot(K, mse_vals, type = "b", xlab = "K", ylab = "Test MSE",
     main = "KNN Regression - Test MSE vs K")
abline(v = best_k, col = "red", lty = 2)
```
The KNN regression model with optimal K = 24 achieved a test MSE of 7.25, indicating an average prediction error of approximately 2.69 grade points on the 0–20 scale. The relatively large optimal K suggests that G3 is best predicted by smoothing across many neighbors, implying a gradual rather than abrupt relationship between the selected predictors and final Portuguese grade.






