---
title: "Logistic_Regression"
author: "Anna Hyunjung Kim"
date: "2026-04-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
student <- read.csv("./student/student-por.csv", header = TRUE, sep = ";")
head(student, 2)
```
```{r}
student <- student[, !(names(student) %in% c("G1", "G2"))]
head(student, 2)
```
```{r}
table(student$G3)
```

```{r}
student[sapply(student, is.character)] <- lapply(student[sapply(student, is.character)], as.factor)
head(student,2)
```

## Target as Pass or Fail. The pass point is 10.
```{r}
student$pass <- ifelse(student$G3 >= 10, 1, 0)
student$pass <- as.factor(student$pass)
```

## Logistic Regression
```{r}
set.seed(321)
train_idx <- sample(1:nrow(student), 0.7*nrow(student))

train <- student[train_idx, ]
test <- student[-train_idx, ]
```

```{r}
# start from full
full_log <- glm(pass ~ . - G3, data = train, family = "binomial")
summary(full_log)
```
```{r}
null_log <- glm(pass ~ 1, data = train, family = "binomial")

step_log <- step(null_log,
                 scope = list(lower = null_log, upper = full_log),
                 direction = "both",
                 trace = FALSE) #- It's too long to show

summary(step_log)
```
The logistic regression results indicate that several variables are significant predictors of student performance. 

In particular, the number of past failures has a strong negative effect on the probability of passing, suggesting that students with previous academic difficulties are less likely to succeed. Similarly, absences also have a negative impact, indicating that increased absenteeism reduces the likelihood of passing.

School type is also significant, with students from MS schools having a lower probability of passing compared to the reference group. In contrast, students who intend to pursue higher education are significantly more likely to pass, reflecting stronger academic motivation.

Gender also appears to play a role, with male students having a lower probability of passing compared to female students. Additionally, students receiving paid classes show a lower probability of passing, which may reflect underlying academic difficulties rather than the direct effect of tutoring.

Some variables, such as father's education (Fedu), are not statistically significant at the 5% level but may still contribute to the overall model based on AIC selection.

## cross validation 
```{r}
library(boot)
```


```{r}
cv_error <- cv.glm(train, step_log, K = 5)
cv_error$delta
```

```{r}
cv_error10 <- cv.glm(train, step_log, K = 10)
cv_error10$delta
```

The cross-validation results show that the model achieves approximately 90% accuracy. The 10-fold cross-validation yields slightly lower error than the 5-fold, indicating more stable performance.

```{r}
log_prob <- predict(step_log, test, type = "response")
log_pred <- ifelse(log_prob > 0.5, 1, 0)
log_pred <- as.factor(log_pred)

mean(log_pred != test$pass)   # error rate
mean(log_pred == test$pass)   # classification rate
```

```{r}
table(test$pass, log_pred)
```

```{r}
table(Actual = test$pass, Predicted = log_pred)
```
# Same process Using a cutoff of 12


### Target as Pass or Fail. The pass point is 12.
```{r}
student$pass <- ifelse(student$G3 >= 12, 1, 0)
student$pass <- as.factor(student$pass)
table(student$pass)
```


### Logistic Regression
```{r}
set.seed(321)
train_idx <- sample(1:nrow(student), 0.7 * nrow(student))

train <- student[train_idx, ]
test <- student[-train_idx, ]
```


```{r}
# Full logistic regression model
full_log <- glm(pass ~ . - G3, data = train, family = "binomial")
summary(full_log)
```


```{r}
# Stepwise logistic regression
null_log <- glm(pass ~ 1, data = train, family = "binomial")

step_log <- step(null_log,
                 scope = list(lower = null_log, upper = full_log),
                 direction = "both",
                 trace = FALSE)

summary(step_log)
```

Some variables with marginal significance (p-value between 0.05 and 0.1) are still included in the model because stepwise selection is based on AIC rather than individual p-values. These variables may contribute to improving overall model fit.

### Cross validation
```{r}
library(boot)
cv_error <- cv.glm(train, step_log, K = 5)
cv_error$delta
cv_error10 <- cv.glm(train, step_log, K = 10)
cv_error10$delta
```

The cross-validation results show that the prediction error is approximately 16.5%, 
corresponding to an accuracy of about 83.5%. 

The results from 5-fold and 10-fold cross-validation are very similar, 
indicating that the model performance is stable.


### Prediction on test set
```{r}
log_prob <- predict(step_log, test, type = "response")
summary(log_prob)
```


```{r}
log_pred <- ifelse(log_prob > 0.5, 1, 0)

actual <- factor(test$pass, levels = c(0, 1))
pred <- factor(log_pred, levels = c(0, 1))

mean(pred != actual)   # error rate
mean(pred == actual)   # classification
```


```{r}
table(Actual = actual, Predicted = pred)
```

### Threshold comparison
```{r}
mean(ifelse(log_prob > 0.3, 1, 0) != as.numeric(as.character(test$pass)))
mean(ifelse(log_prob > 0.7, 1, 0) != as.numeric(as.character(test$pass)))
```