---
title: "FinalProject"
author: "Aditya Kataria"
date: "2026-04-08"
output: html_document
---

```{r}
library(tidyverse)
library(MASS)
```
# load data
```{r}
student <- read.csv("./student/student-por.csv", header = TRUE, sep = ";")
```

# overview of data
```{r}
head(student)
str(student)
dim(student) 
```

# convert characters to factors
```{r}
student[sapply(student, is.character)] <- lapply(student[sapply(student, is.character)], as.factor)
```


# Logistic regression
```{r}
# Full model
full_model <- lm(G3 ~ . - G1 - G2, data = student)

# Null model 
null_model <- lm(G3 ~ 1, data = student)

# AIC stepwise
step_aic <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(step_aic)

# BIC stepwise
step_bic <- stepAIC(full_model, direction = "both", k = log(nrow(student)), trace = FALSE)
summary(step_bic)

# Compare number of predictors selected
length(coef(step_aic)) - 1  # AIC selected predictors
length(coef(step_bic)) - 1  # BIC selected predictors

# Compare AIC and BIC values
AIC(step_aic)
BIC(step_bic)
```
To perform variable selection, stepwise regression was applied in both directions (forward and backward) using AIC and BIC as selection criteria on a model predicting Portuguese final grade (G3) from all available predictors excluding prior period grades G1 and G2. BIC applies a heavier penalty for model complexity, so it tends to produce more sparse models.

The AIC-selected model retained 14 predictors: school, sex, age, mother's education (Medu), guardian, study time, failures, school support, higher education aspiration, romantic relationship, weekday alcohol consumption (Dalc), health, and absences. The BIC-selected model retained 8 predictors: school, father's education (Fedu), study time, failures, school support, higher education aspiration, Dalc, and health. Both models were statistically significant (p < 2.2e-16).

The most substantive predictors in both models were past failures (B = -1.48, p < 0.001), which reduced G3 by nearly 1.5 points per failure, and higher education aspiration (B = +1.86, p < 0.001), which shows that motivation has a strong effect on grades. Students at MS school scored approximately 1.5 points lower than GP students. School support showed a negative coefficient (B = -1.33), which reflects selection bias: struggling students are more likely to receive extra support. Parental education (Medu in AIC, Fedu in BIC) had a modest positive effect, and weekday alcohol consumption (Dalc) negatively impacted performance in both models.



# cross validation 
```{r}
library(boot)

# AIC model
aic_model <- glm(G3 ~ school + sex + age + Medu + guardian + studytime + 
                   failures + schoolsup + higher + romantic + Dalc + health + 
                   absences, data = student)

# BIC model
bic_model <- glm(G3 ~ school + Fedu + studytime + failures + schoolsup + 
                   higher + Dalc + health, data = student)

# 10-fold CV for both
set.seed(42)
cv_aic <- cv.glm(student, aic_model, K = 10)
cv_bic <- cv.glm(student, bic_model, K = 10)

# CV MSE
cv_aic$delta[1]  # AIC model CV MSE
cv_bic$delta[1]  # BIC model CV MSE
```
10-fold cross-validation yielded a CV MSE of 7.42 for the AIC model and 7.40 for the BIC model, corresponding to average prediction errors of approximately 2.72 grade points on the 0–20 scale. Despite using 6 fewer predictors, the BIC model matched the AIC model's out-of-sample accuracy, making it the preferred stepwise result on grounds of parsimony. Both models explain approximately 31-33% of variance in G3, which is expected for a purely demographic and behavioral model without prior grade information.





